---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# dremeR

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
<!-- badges: end -->

An R interface to the [MEME Suite](http://meme-suite.org/) family of tools.

## Installation


```{r, include=F}
#You can install the released version of dremeR from [CRAN](https://CRAN.R-project.org) with:
install.packages("dremeR")
#And the development version from [GitHub](https://github.com/) with:
```

You can install the development version of dremeR from [GitHub](https://github.com/) with:

```{r, eval=F}
# install.packages("remotes")
remotes::install_github("snystrom/dotargs")
remotes::install_github("snystrom/dremeR")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(dremeR)
## basic example code
```

```{r}
fa <- "inst/extdata/fasta_ex/fa1.fa"
dreme_out <- runDreme(fa, "shuffle", e = 39)
```

dreme results are a `data.frame`. The `motifs` column contains a `universalmotif` object with the PCM information for each *de-novo* discovered motif. This is so that any filtering of the results object also simply filter the available motifs.
```{r}
dreme_out
```
```{r}
library(universalmotif)

view_motifs(dreme_out$motifs)
```
The primary advantage of using the `data.frame` output allows simple integration with base subsetting, piping, and the `tidyverse`.
```{r}
library(dplyr)

dreme_out %>% 
  filter(length == 3) %>% 
  {universalmotif::view_motifs(.$motifs)}
```

`universalmotif` manipulations can easily be executed on data.frame columns. For example:
```{r, fig.height=1.5}
dreme_out$motifs %>% 
  merge_motifs() %>% 
  view_motifs()
```

