---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
  #out.width = "100%"
)
```

# memes

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
<!-- badges: end -->

An R interface to the [MEME Suite](http://meme-suite.org/) family of tools,
which provides several utilities for performing motif analysis on DNA, RNA, and
protein sequences. memes works by detecting a local install of the MEME suite,
running the commands, then importing the results directly into R.

## Installation

### Development Version
You can install the development version of memes from [GitHub](https://github.com/snystrom/memes) with:

```{r, eval=F}
if (!requireNamespace("remotes", quietly=TRUE))
  install.packages("remotes")
# memes currently requires the development version of universalmotif:
remotes::install_github("bjmt/universalmotif")
remotes::install_github("snystrom/memes")
```


## Detecting the MEME Suite

memes relies on a local install of the [MEME Suite](http://meme-suite.org/).
For installation instructions for the MEME suite, see the [MEME Suite
Installation Guide](http://meme-suite.org/doc/install.html?man_type=web).

memes needs to know the location of the `meme/bin/` directory on your local machine.
You can tell memes the location of your MEME suite install in 4 ways. memes
will always prefer the more specific definition if it is a valid path. Here they
are ranked from most- to least-specific:

1. Manually passing the install path to the `meme_path` argument of all memes functions
2. Setting the path using `options(meme_bin = "/path/to/meme/bin/")` inside your R script
3. Setting `MEME_BIN=/path/to/meme/bin/` in your `.Renviron` file
4. memes will try the default MEME install location `~/meme/bin/`

If memes fails to detect your install at the specified location, it will fall
back to the next option.

To verify memes can detect your MEME install, use `check_meme_install()` which
uses the search herirarchy above to find a valid MEME install. It will report
whether any tools are missing, and print the path to MEME that it sees. This can
be useful for troubleshooting issues with your install.
```{r check_install_works}
library(memes)

# Verify that memes detects your meme install
# (returns all green checks if so)
# (I have MEME installed to the default location)
check_meme_install()
```

```{r check_install_fails}
# You can manually input a path to meme_path
# If no meme/bin is detected, will return a red X
check_meme_install(meme_path = 'bad/path')
```

## The Core Tools

| Function Name | Use              | Sequence Input | Motif Input | Output |
|:-------------:|:----------------:|:--------------:|:-----------:|:-------------------------------------------------------|
| `runDreme()`  | Motif Discovery (short motifs)  | Yes | No      | data.frame w/ `motifs` column                          |
| `runAme()`    | Motif Enrichment                | Yes | Yes     | data.frame (optional: `sequences` column)              |
| `runFimo()`   | Motif Scanning                  | Yes | Yes     | GRanges of motif positions                             |
| `runTomTom()` | Motif Comparison                | No  | Yes     | data.frame w/ `best_match_motif` and `tomtom` columns* |
| `runMeme()`   | Motif Discovery (long motifs)   | Yes | No      | data.frame w/ `motifs` column                          |

\* **Note:** if `runTomTom()` is run using the results of `runDreme()`, the results
will be joined with the `runDreme()` results as extra columns. This allows easy
comparison of *de-novo* discovered motifs with their matches.

**Sequence Inputs** can be any of:

1. Path to a .fasta formatted file
2. `Biostrings::XStringSet` (can be generated from GRanges using `get_sequence()` helper function)
3. A named list of `Biostrings::XStringSet` objects (generated by `get_sequence()`)

**Motif Inputs** can be any of:

1. A path to a .meme formatted file of motifs to scan against
2. A `universalmotif` object, or list of `universalmotif` objects
3. A `runDreme()` results object (this allows the results of `runDreme()` to pass directly to `runTomTom()`)
4. A combination of all of the above passed as a `list()` (e.g. `list("path/to/database.meme", "dreme_results" = dreme_res)`)

**Output Types**:

`runDreme()`, and `runTomTom()` return data.frames with special columns. The `motif`
column contains `universalmotif` objects, with 1 entry per row. The remaining
columns describe the properties of each returned motif. The following column
names are special in that their values are used when running `update_motifs` to
alter the properties of the motifs stored in the `motif` column. Be careful
about changing these values.

 - name
 - altname
 - family
 - organism
 - strand
 - nsites
 - bkgsites
 - pval
 - qval
 - eval

memes is built around the [universalmotif package](https://www.bioconductor.org/packages/release/bioc/html/universalmotif.html)
which provides a framework for manipulating motifs in R. The `motif` columns
from `runDreme()` and `runTomTom()` can be used natively with all `universalmotif`
functions (eg `universalmotif::view_motifs(dreme_results$motif)`).

`runTomTom()` returns a special column: `tomtom` which is a `data.frame` of all match data for each input motif. This can be expanded out using `tidyr::unnest(tomtom_results, "tomtom")`.

## Quick Examples
### Motif Discovery with DREME

```{r}
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(GenomicRanges))

# Example transcription factor peaks as GRanges
data("example_peaks", package = "memes")

# Genome object
dm.genome <- BSgenome.Dmelanogaster.UCSC.dm6::BSgenome.Dmelanogaster.UCSC.dm6
```
The `get_sequence` function takes a `GRanges` or `GRangesList` as input and
returns the sequences as a `BioStrings::XStringSet`, or list of `XStringSet`
objects, respectively. `get_sequence` will name each fasta entry by the genomic
coordinates each sequence is from.
```{r}
# Generate sequences from 200bp about the center of my peaks of interest
sequences <- example_peaks %>% 
  resize(200, "center") %>% 
  get_sequence(dm.genome)
```

`runDreme()` accepts XStringSet or a path to a fasta file as input. You can use
other sequences or shuffled input sequences as the control dataset.
```{r}
# runDreme accepts all arguments that the commandline version of dreme accepts
# here I set e = 50 to detect motifs in the limited example peak list
dreme_results <- runDreme(sequences, control = "shuffle", e = 50)
```
memes is built around the
[universalmotif](https://www.bioconductor.org/packages/release/bioc/html/universalmotif.html)
package. The `motif` column of the runDreme results object contains
`universalmotif` objects which can be used natively in all `universalmotif`
functions.
```{r}
library(universalmotif)
view_motifs(dreme_results$motif)
```

### Matching motifs using TOMTOM
Discovered motifs can be matched to known TF motifs using `runTomTom()`, which can accept as input a path to a .meme formatted file, a `universalmotif` list, or the results of `runDreme()`.

TomTom uses a database of known motifs which can be passed to the `database` parameter as a path to a .meme format file, or a `universalmotif` object.

Optionally, you can set the environment variable `MEME_DB` in `.Renviron`, or
the `meme_db` value in `options` to a valid .meme format file and memes will
use that file as the database. memes will always prefer user input to the
function call.

```{r}
options(meme_db = system.file("extdata/db/fly_factor_survey_id.meme", package = "memes"))
m <- create_motif("CMATTACN", altname = "testMotif")
tomtom_results <- runTomTom(m)
```
```{r}
tomtom_results
```

### Using runDreme results as TOMTOM input
`runTomTom()` will add its results as columns to a `runDreme()` results data.frame.
```{r}
full_results <- dreme_results %>% 
  runTomTom()
```


### Motif Enrichment using AME

AME is used to test for enrichment of known motifs in target sequences. `runAme()`
will use the `MEME_DB` entry in `.Renviron` or `options(meme_db =
"path/to/database.meme")` as the motif database. Alternately, it will accept all
valid inputs as `runTomTom()`.
```{r}
ame_results <- runAme(sequences, control = "shuffle", evalue_report_threshold = 30)
ame_results
```


## Visualizing Results

`view_tomtom_hits` allows comparing the input motifs to the top hits from
TomTom. Manual inspection of these matches is important, as sometimes the top
match is not always the correct assignment. Altering `top_n` allows you to show
additional matches in descending order of their rank.

```{r}
full_results %>% 
  view_tomtom_hits(top_n = 1) %>% 
  # Here I use the cowplot library simply to combine each plot into a 4-panel grid
  cowplot::plot_grid(plotlist = .)
```

It can be useful to view the results from `runAme()` as a heatmap. 
`ame_plot_heatmap()` can create complex visualizations for analysis of enrichment
between different region types (see vignettes for details). Here is a simple
example heatmap.
```{r, fig.height=3, fig.width=5}
ame_results %>% 
  ame_plot_heatmap()
```

## Importing Data from previous runs

memes also supports importing results generated using the MEME suite outside of R (for example, running jobs on [meme-suite.org](meme-suite.org), or running on the commandline). 

| MEME Tool | Function Name       | File Type        |
|:---------:|:-------------------:|:----------------:|
| Dreme     | `importDremeXML()`  | dreme.xml        |
| TomTom    | `importTomTomXML()` | tomtom.xml       |
| AME       | `importAme()`       | ame.tsv*         |
| FIMO      | `importFimo()`      | fimo.tsv         | 
| Meme      | `importMeme()`      | meme.txt         | 

* `importAME()` can also use the "sequences.tsv" output when AME used `method = "fisher"`, this is optional.


# FAQs
### How do I use memes/MEME on Windows?
The MEME Suite does not currently support Windows, although it can be
installed under [Cygwin](https://www.cygwin.com/) or the [Windows Linux Subsytem](https://docs.microsoft.com/en-us/windows/wsl/install-win10) (WSL).
Please note that if MEME is installed on Cygwin or WSL, you must also run R
inside Cygwin or WSL to use memes.

# Citation

memes is a wrapper for a select few tools from the MEME Suite, which were
developed by another group. In addition to citing memes, please cite the MEME
Suite tools corresponding to the tools you use.

If you use `runDreme()` in your analysis, please cite:

Timothy L. Bailey, "DREME: Motif discovery in transcription factor ChIP-seq data", Bioinformatics, 27(12):1653-1659, 2011. [full text](https://academic.oup.com/bioinformatics/article/27/12/1653/257754)

If you use `runTomTom()` in your analysis, please cite:

Shobhit Gupta, JA Stamatoyannopolous, Timothy Bailey and William Stafford Noble, "Quantifying similarity between motifs", Genome Biology, 8(2):R24, 2007. [full text](http://genomebiology.com/2007/8/2/R24)

If you use `runAme()` in your analysis, please cite:

Robert McLeay and Timothy L. Bailey, "Motif Enrichment Analysis: A unified framework and method evaluation", BMC Bioinformatics, 11:165, 2010, doi:10.1186/1471-2105-11-165. [full text](http://www.biomedcentral.com/1471-2105/11/165)

If you use `runFimo()` in your analysis, please cite:

Charles E. Grant, Timothy L. Bailey, and William Stafford Noble, "FIMO: Scanning for occurrences of a given motif", Bioinformatics, 27(7):1017-1018, 2011. [full text](http://bioinformatics.oxfordjournals.org/content/early/2011/02/16/bioinformatics.btr064.full)

## Licensing Restrictions
The MEME Suite is free for non-profit use, but for-profit users should purchase a
license. See the [MEME Suite Copyright Page](http://meme-suite.org/doc/copyright.html) for details.
