---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
  #out.width = "100%"
)
```

# dremeR

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
<!-- badges: end -->

An R interface to the [MEME Suite](http://meme-suite.org/) family of tools.

## Installation


```{r, include=F, eval=F}
#You can install the released version of dremeR from [CRAN](https://CRAN.R-project.org) with:
install.packages("dremeR")
#And the development version from [GitHub](https://github.com/) with:
```

You can install the development version of dremeR from [GitHub](https://github.com/) with:

```{r, eval=F}
# install.packages("remotes")
remotes::install_github("snystrom/dotargs")
remotes::install_github("snystrom/dremeR")
```
## Example

```{r example}
library(dremeR)

# Verify that dremeR detects your meme install
# should return all green checks if so.
check_meme_install()
```

```{r}
fa <- dremeR:::duplicate_file("inst/extdata/fasta_ex/fa1.fa")
dreme_out <- runDreme(fa, "shuffle", e = 39)
```

dreme results are a `data.frame`. The `motif` column contains a `universalmotif` object with the PCM information for each *de-novo* discovered motif. This is so that any filtering of the results object also simply filter the available motifs.
```{r}
dreme_out
```
```{r}
library(universalmotif)

view_motifs(dreme_out$motif)
```
The primary advantage of using the `data.frame` output allows simple integration with base subsetting, piping, and the `tidyverse`.
```{r}
library(dplyr)

dreme_out %>% 
  filter(length == 3) %>% 
  {universalmotif::view_motifs(.$motif)}
```

`universalmotif` manipulations can easily be executed on data.frame columns. For example:
```{r, fig.height=1.5}
dreme_out$motif %>% 
  merge_motifs() %>% 
  view_motifs()
```

## TomTom 
TomTom can be used to match *de-novo* motifs from dreme to a database of known
motifs. If run on the output of `runDreme` it will append information for the best match in columns prefixed `best_`. Secondary matches found by tomtom will be stored in the list column `tomtom`. 
```{r}
full_res <- runTomTom(dreme_out, database = "inst/extdata/db/fly_factor_survey_id.meme")
```
```{r}
full_res
```
```{r}
full_res$tomtom
```

```{r}
full_res %>% 
  dplyr::select(id, alt, seq, best_match_id, best_match_pvalue, best_match_evalue, best_match_motif)
```
```{r}
full_res$best_match_motif

view_tomtom_hits(full_res, 1)
```

## Denovo Motif Pipeline w/ dremer
```{r, message=F}
suppressPackageStartupMessages(library(GenomicRanges))

peaks <- "inst/extdata/peaks/peaks.tsv" %>%
  readr::read_tsv() %>%
  GRanges
```
```{r}
dm.genome <- BSgenome.Dmelanogaster.UCSC.dm6::BSgenome.Dmelanogaster.UCSC.dm6
options(meme_db = "inst/extdata/db/fly_factor_survey_id.meme")

motif_analysis <- peaks %>% 
  resize(200, "center") %>% 
  get_sequence(dm.genome) %>% 
  runDreme("shuffle", e = 50) %>% 
  runTomTom()
```

## Compare denovo motif with best match

use `cowplot` to arrange plots in a grid.
```{r}
view_tomtom_hits(motif_analysis, 1) %>% 
  cowplot::plot_grid(plotlist = ., labels = "AUTO")
```

## Evaluate top 3 hits from TOMTOM

```{r}
view_tomtom_hits(motif_analysis, 3)
```
```{r}
library(ggplot2)

frac_plot <- motif_analysis %>% 
  ggplot(aes(reorder(id, pos_frac), pos_frac/neg_frac)) +
    geom_col() + 
    coord_flip() +
    labs(x = NULL,
         y = "Enrichment Ratio") +
    theme_bw() +
    theme(axis.text.y = element_blank())
```

```{r}
cowplot::plot_grid(
  universalmotif::view_motifs(motif_analysis$motif),
  frac_plot
)
```

## Motif scanning & enrichment testing with AME
```{r}
ame_analysis <- peaks %>% 
  resize(200, "center") %>% 
  get_sequence(dm.genome) %>% 
  runAme(evalue_report_threshold = 50)
```
```{r}
ame_analysis %>% 
  ame_plot_heatmap()
```



### Example for converting gene ids
Try using `annotationDbi::mapIds` instead. Also, note below E93 Fbgn is out of date & does not map.
```{r}
ame_analysis$motif_alt_id %>% 
  gsub("_.+", "", .) %>% 
  clusterProfiler::bitr(fromType = "FLYBASE", toType = "SYMBOL", OrgDb = org.Dm.eg.db::org.Dm.eg.db)
```

### reducing best matches from ame
```{r}

ame_analysis %>% 
  dplyr::group_by(motif_id) %>% 
  dplyr::filter(adj.pvalue == min(adj.pvalue)) %>% 
  dplyr::ungroup()
  
```

### Ame in partitioning mode
```{r}
ame_partition <- peaks %>% 
  data.frame %>% 
  # NOTE: never actually do this, just a quick way to generate example data
  dplyr::mutate(score = seq_along(start)) %>% 
  GRanges %>% 
  resize(200, "center") %>% 
  get_sequence(dm.genome, score = "score") %>% 
  runAme(control = NA, evalue_report_threshold = 50)
```
```{r, fig.height=3, fig.width=10}
ame_partition %>% 
  dplyr::mutate(tfid = gsub("(.+)_{1}.+", "\\1", motif_id)) %>% 
  ame_plot_heatmap(id = tfid)
```

### Dotargs backend & error checking

```{r, eval=F}
ame_analysis <- peaks %>% 
  resize(200, "center") %>% 
  get_sequence(dm.genome) %>% 
  runAme(evalue_reportt_threshold = 30)
```
```{r, echo = F, error = T}
usethis::ui_stop("Invalid flags. Did you mean: \"evalue_report_threshold\" instead of: \"evalue_reportt_threshold\"")
```

