---
title: "Install MEME"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Install MEME}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

dremeR is an R interface to the [MEME Suite](http://meme-suite.org/) family of tools,
which provides several utilities for performing motif analysis on DNA, RNA, and
protein sequences. It works by detecting a local install of the MEME suite,
running the commands, then importing the results directly into R.

## Installing the MEME Suite

dremeR relies on a local install of the [MEME Suite](http://meme-suite.org/).
For installation instructions for the MEME suite, see the [MEME Suite Installation Guide](http://meme-suite.org/doc/install.html?man_type=web).

Briefly, the MEME suite can be installed to a default location (`~/meme/`) on
Linux, MacOS, Cygwin, and Windows Linux Subsystem using the following shell
commands:

```{bash, eval=F}
# As of May 2020, version 5.1.1 is the most recent MEME-Suite version
# Please check the install guide (linked above) for more recent information
version=5.1.1
wget http://meme-suite.org/meme-software/$version/meme-$version.tar.gz
tar zxf meme-$version.tar.gz
cd meme-$version
./configure --prefix=$HOME/meme --with-url=http://meme-suite.org/ --enable-build-libxml2 --enable-build-libxslt
make
make test
make install
```

For additional troubleshooting or to learn more about install configuration, please see the [Installation Guide](http://meme-suite.org/doc/install.html?man_type=web).

## Detecting the MEME Suite

dremeR needs to know the location of the `meme/bin/` directory on your local machine.
You can tell dremeR the location of your MEME suite install in 4 ways. dremeR
will always prefer the more specific definition if it is a valid path. Here they
are ranked from most- to least-specific:

1. Manually passing the install path to the `meme_path` argument of all dremeR functions
2. Setting the path using `options(meme_bin = "/path/to/meme/bin/")` inside your R script
3. Setting `MEME_BIN=/path/to/meme/bin/` in your `.Renviron` file
4. dremeR will try the default MEME install location `~/meme/bin/`

If dremeR fails to detect your install at the specified location, it will fall
back to the next option.

To verify dremeR can detect your MEME install, use `check_meme_install()` which
uses the search herirarchy above to find a valid MEME install. It will report
whether any tools are missing, and print the path to MEME that it sees. This can
be useful for troubleshooting issues with your install.

```{r check_install_works}
library(dremeR)

# Verify that dremeR detects your meme install
# (returns all green checks if so)
# (I have MEME installed to the default location)
check_meme_install()
```

```{r check_install_fails}
# You can manually input a path to meme_path
# If no meme/bin is detected, will return a red X
check_meme_install(meme_path = 'bad/path')
```
