---
title: "Integrative Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integrative Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dremeR)
library(magrittr)
suppressPackageStartupMessages(library(GenomicRanges))
```

```{r}
check_meme_install()
```


```{r, message=F}
# TODO: Change this to use system.file(extadata/etc.) once example data is finalized.
#"../inst/extdata/peaks/e93_chr3.csv"
#system.file("extdata/peaks/e93_chr3.csv")
warning("Remember to update the correct file path")
peaks <- "../inst/extdata/peaks/e93_chr3.csv" %>% 
  readr::read_csv() %>% 
  GRanges
```

### Motifs in opening vs closing sites
```{r}
# These data use the dm3 reference genome
dm.genome <- BSgenome.Dmelanogaster.UCSC.dm3::BSgenome.Dmelanogaster.UCSC.dm3

# compute summits using the summit_position column
summits <- peaks %>% 
  plyranges::anchor_start() %>% 
  plyranges::mutate(width = 1) %>% 
  plyranges::shift_right(mcols(.)$summit_position) 
  
# Get sequences in a 100bp window around the peak summit
summit_flank <- summits %>% 
  plyranges::anchor_center() %>% 
  plyranges::mutate(width = 100) 

# split by response to E93 binding
by_sens <- summit_flank %>% 
  split(mcols(.)$e93_sensitive_behavior) %>% 
  get_sequence(dm.genome)
```

```{r dreme_by_sens_vs_static, cache=TRUE, cache.comments=FALSE}
dreme_by_sens_vs_static <- by_sens[c("Increasing", "Decreasing")] %>% 
  lapply(function(seq){
    runDreme(seq, control = by_sens$Static)
  }) 
```

```{r}
dreme_results <- dreme_by_sens_vs_static %>% 
  dplyr::bind_rows(.id = "e93_response")
```
```{r}
#options(meme_db = )
options(meme_db = "../inst/extdata/db/fly_factor_survey_id.meme")
all_res <- dreme_results %>% 
  runTomTom(dist = "ed")
```

```{r, fig.height=8, fig.width=8}
all_res %>% 
  view_tomtom_hits(3) %>% 
  cowplot::plot_grid(plotlist = ., labels = "AUTO")
```
```{r}
all_res %<>% 
  force_best_match(c("m01_CSAAAAM" = "Eip93F_SANGER_10",
                     "m03_AKGG" = "pho_SANGER_10"))
```


```{r, fig.height=4.5, fig.width=8}
all_res %>% 
  view_tomtom_hits(1) %>% 
  cowplot::plot_grid(plotlist = ., labels = "AUTO")
```

```{r, fig.height=3, fig.width=8}
all_res %>% 
  dplyr::mutate(label = paste0(e93_response, " in response to E93")) %>% 
  split(.$label) %>% 
  purrr::imap(~{
    top_hits <- view_tomtom_hits(.x, 1) %>% 
      cowplot::plot_grid(plotlist = ., nrow = 1, labels = "AUTO")
    
    title <- cowplot::ggdraw() +
        cowplot::draw_text(.y)
    
    cowplot::plot_grid(plotlist = list(title, top_hits), 
                       ncol = 1,
                       rel_heights = c(0.1, 1)
    )
  })
```

### Motif scanning with AME
```{r}
by_binding <- summit_flank %>% 
  split(mcols(.)$peak_binding_description) %>% 
  get_sequence(dm.genome)
```
```{r}
ame_by_binding <- by_binding %>% 
  purrr::map(runAme) 
```

### Visualizing AME results

The `ame_plot_heatmap()` function provides a quick way to visualize AME results.
It is built on top of `ggplot2`, so all `ggplot2` functions can be used to
further modify the plot.

By default, it uses the -log10(adjusted p-value) as the heat values. See the
documentation for additional notes on customization.
```{r, fig.height=3.5, fig.width=7}
library(ggplot2)

ame_by_binding$entopic %>% 
  dplyr::filter(rank %in% 1:10) %>% 
  ame_plot_heatmap(group_name = "Entopic Sites")  +
    ggtitle("Top 10 AME Hits in Entopic Sites")
```

To plot results from multiple runs together, they must first be joined into 1
data frame. The `ame_by_binding` object is a list whose names correspond to the
E93 binding category. The list can be combined into a data.frame using
`dplyr::bind_rows`. Setting `.id = "binding_type` creates a new column
`binding_type` that contains the names from the `ame_by_binding` list. In this
way, the `ame_res` data.frame contains all AME results for each run, which can
be distinguished by the `binding_type` column.
```{r}
ame_res <- ame_by_binding %>% 
  dplyr::bind_rows(.id = "binding_type")
  
```

It is possible to aggregate results from multiple runs into a heatmap by setting
the `group` parameter in `ame_plot_heatmap()`.

This is too many hits to view in this vignette, but you can see that the heatmap
will plot motifs by their overlap across groups, where unique motifs are on the
left, and shared motifs are on the right.
```{r, fig.height=5,fig.width=15}
ame_res %>% 
  ame_plot_heatmap(group = binding_type)
```

### Reducing redundant motif hits
There is another key consideration for the above visualization and that is that
in the FlyFactorSurvey database we used, different TFs can have multiple motif
entries in the database which are all detected separately by AME. Here, when
returning the top 5 hits from each group, you can see, for example, that a motif
matching "Aef1" is reported 3 times within the top 5 hits of the ectopic sites
(A similar result can also be seen in the top 10 entopic hits heatmap above).
In this situation, it makes sense to summarize the data at the TF level, instead
of the motif level. **Note:** There may be exceptions to this if for example, a
TF has multiple DNA binding sequences it can recognize, in which case having
multiple hits may reflect a biological property of your sequences. You will have
to handle this on a case-by-case basis for interesting hits and different motif
databases. Here, we can see that at least for Aef1, the consensus sequences are
very similar.
```{r}
ame_res %>% 
  dplyr::group_by(binding_type) %>% 
  dplyr::filter(rank %in% 1:5) %>% 
  head(5) %>% 
  dplyr::select(binding_type, rank, motif_id, motif_alt_id, consensus)
```

How to solve this problem will vary with different motif databases (For details
on how to pre-process a motif database, see `vignette("tidy_motifs")`). In the
FlyFactorSurvey database, the `motif_alt_id` column contains a unique FBgn
number for each TF. For entries of the same TF, the FBgn number is prepended
with an `_<number>`. We can extract the FBgn number to use as a gene-level
identifier, and reduce the matches to return 1 match per gene as follows:
```{r, fig.height=3, fig.width=25}
ame_res %>% 
  # the flyfactor genes have their Flybase gene numbers appended with a _<number> 
  # this is to make the entries unique, but the gene ID is encoded before the _
  # so I remove the _<number> with a regex, leaving the FBgn number.
  dplyr::mutate(gene_id = gsub("_\\d+", "", motif_alt_id)) %>% 
  dplyr::ungroup() %>% 
  # Reduce redundant TF hits by returning only the match with the lowest p-value
  # for each TF. It is important to group by binding_type because the analyses
  # were run separately, so these are different results (we want to return the
  # lowest p-value for each group a motif was found within).
  dplyr::group_by(binding_type, gene_id) %>% 
  dplyr::filter(adj.pvalue == min(adj.pvalue)) %>% 
  # add new id column to simplify motif name to TF name
  dplyr::mutate(tfid = strsplit(motif_id, "_")[[1]][1]) %>% 
  ame_plot_heatmap(group = binding_type, id = tfid)
```


## Issues with AME Heatmap Visualization
Finally, because the data for each binding type are generated from different AME
runs, it is not the best idea to compare p-values across experiments (ie between
rows), because the dynamic range of p-values will vary between groups. Here you
can see that, for example, the orphan sites will have a very different scale of
-log10(adjusted pvalues) than ectopic and entopic sites. In other words, because
the dynamic range of values are different between experiments, placing them on
the same scale for comparison may be misleading in certain contexts.

```{r}
ame_res %>% 
  ggplot(aes(-log10(adj.pvalue))) +
    stat_ecdf(aes(color = binding_type), size = 1, pad = FALSE) +
    labs(y = "Fraction of Motifs") +
    theme_bw()
```

We can partially overcome this limitation by filling the heatmap with the
normalized rank value for each TF, which accounts for differences in total number of
discovered motifs between AME runs. Although it does not completely abrograte
differeces, the signal values for high-ranked motifs within groups will be more
comparable.
```{r}
# rank_normalize transforms the ranks to values between 0 and 1 as follows:
# (1 - rank)/(1 - max(rank))
ame_res %>% 
  ggplot(aes(dremeR:::rank_normalize(rank))) +
    stat_ecdf(aes(color = binding_type), size = 1, pad = FALSE) +
    scale_x_continuous(
                       breaks = c(0, 0.25, 0.5, 0.75, 1),
                       labels = c("0\n(High)", 0.25, 0.5, 0.75, "1\n(Low)"),
                       ) +
    labs(y = "Fraction of Motifs",
         x = "Normalized Rank")  +
    theme_bw() 
```

We achieve this by setting `value = "normalize"` in `ame_plot_heatmap()`. If
plotting with multiple groups, it may be more appropriate to make this
visualization.

```{r, fig.height=3, fig.width=25}
ame_res %>% 
  dplyr::mutate(gene_id = gsub("_\\d+", "", motif_alt_id)) %>% 
  dplyr::ungroup() %>% 
  dplyr::group_by(binding_type, gene_id) %>% 
  dplyr::filter(adj.pvalue == min(adj.pvalue)) %>% 
  dplyr::mutate(tfid = strsplit(motif_id, "_")[[1]][1]) %>% 
  ame_plot_heatmap(group = binding_type, id = tfid, value = "normalize")
```


## FIMO 
Because we discover a *de-novo* motif from DREME that matches to E93, it can be useful to scan peaks for motif occurrances to further examine properties of different peak categories.

Although we could also import the motifs from our local .meme database file, it is also possible to use motifs pulled from a [`MotifDb`](https://bioconductor.org/packages/release/bioc/html/MotifDb.html) query as follows. 
```{r}
e93_motif <- MotifDb::MotifDb %>% 
  MotifDb::query("Eip93F") %>% 
  universalmotif::convert_motifs() %>% 
  .[[1]]

# Rename the motif for simplicity
e93_motif["name"] <- "E93"

e93_motif
```

```{r}
fimo_res <- summit_flank %>% 
  get_sequence(dm.genome) %>% 
  runFimo(motifs = e93_motif, thresh = 1e-3)
```

```{r}
fimo_res
```

Here we add a count for each motif per peak in the `n_motifs` column. We also add a column `has_motif` which will be a binary indicator of whether a peak contains any motifs.
```{r}
summit_flank %<>% 
  # Note, if running FIMO with multiple motifs, this solution will not work
  # as it will count all motifs within the fimo-results without splitting by motif_id
  plyranges::mutate(n_motifs = plyranges::count_overlaps(., fimo_res), 
                    has_motif = n_motifs > 0)
```

Ask whether E93 sensitive sites are more likely to have E93 motifs in certain response types. Here we can see that sensitive decreasing sites are more likely to have E93 motifs than sensitive increasing or insensitive static sites.
```{r, fig.height=4, fig.width=5}
summit_flank %>% 
  data.frame %>% 
  dplyr::mutate(has_match = dplyr::if_else(has_motif, "Match", "No Match")) %>% 
  ggplot(aes(e93_sensitive_behavior)) +
    geom_bar(aes(fill = forcats::fct_rev(has_match)), position = "fill") +
    scale_fill_manual(values = c("Match" = "firebrick",
                                 "No Match" = "Black")) +
    labs(fill = "E93 Motif Match")
```


To investigate whether E93-sensitive sites that open/close/don't change have
different numbers of E93 motifs, we plot the fraction of sites with each number
of motifs. Here we can see that in addition to being more likely to contain an E93 motif, sensitive decreasing sites are more likely to to contain 2 or more matches, where 10% contain at least 2 motifs.
```{r, fig.height=5, fig.width=7}
summit_flank %>% 
  # currently, group operations are faster as data.frames, so we convert to data.frame
  data.frame %>% 
  dplyr::group_by(e93_sensitive_behavior, n_motifs) %>% 
  dplyr::count() %>% 
  dplyr::group_by(e93_sensitive_behavior) %>% 
  dplyr::mutate(frac = n/sum(n)) %>% 
  ggplot(aes(n_motifs, frac)) +
    geom_line(aes(color = e93_sensitive_behavior), size = 1) +
    labs(y = "Fraction of Sites",
         x = "Number of E93 Motifs") +
    theme_bw()
```

Finally, we want to assess whether the quality of E93 motifs is different between sensitivity categories. To examine this, we need to determine which motifs are found in which peaks. We use the `plyranges::join_overlap_intersect` to return motif entries appended with peak-level metadata, like the peak id each motif is found within.
```{r}
# return position of each motif match w/ peak meatadata
intersect <- summit_flank %>% 
  plyranges::join_overlap_intersect(fimo_res)
```

We use the FIMO `score` as a proxy for quality, where higher scores are better
matches to the motif. Here we examine only the best match (highest score) motif
within each peak. 

```{r, fig.height=4, fig.width=3}
intersect %>% 
  data.frame %>% 
  dplyr::group_by(id) %>% 
  dplyr::filter(score == max(score)) %>% 
  ggplot(aes(e93_sensitive_behavior, score)) +
    geom_boxplot(aes(fill = e93_sensitive_behavior), notch = T, width = 0.5) +
    guides(fill = "none") +
    labs(x = "Response to E93",
         y = "FIMO Score") +
    theme_bw()
  
```
```{r}
intersect %>% 
  data.frame %>% 
  dplyr::group_by(id) %>% 
  dplyr::filter(score == max(score)) %>% 
  dplyr::mutate(behavior = factor(e93_sensitive_behavior)) %>% 
  PMCMRplus::adAllPairsTest(score ~ behavior, ., p.adjust.method = "fdr")
```


### Centrality of E93 motif
Next we want to visualize whether the E93 motif has altered positioning across the response categories. To do this we 
```{r}
distance_to_nearest <- function(query, subject, ...){
  hits <- GenomicRanges::distanceToNearest(query, subject, ...)
  mcols(hits)$distance
}

summit_nearest_e93_motif <- summit_flank %>% 
  plyranges::anchor_center() %>% 
  plyranges::mutate(width = 1) %>% 
  plyranges::join_nearest(fimo_res) %>% 
  plyranges::mutate(dist_to_motif = distance_to_nearest(., fimo_res))

```

```{r}
summit_nearest_e93_motif %>% 
  data.frame %>% 
  dplyr::filter(has_motif == TRUE) %>% 
  ggplot(aes(dist_to_motif)) +
    stat_ecdf(aes(color = e93_sensitive_behavior), size = 1, pad = FALSE) +
    labs(x = "Distance to nearest E93 Motif",
         y = "Fraction of Sites",
         color = "E93 Response") +
    theme_linedraw() 
  
```
```{r}
summit_nearest_e93_motif %>% 
  data.frame %>% 
  dplyr::filter(has_motif == TRUE) %>% 
  dplyr::mutate(behavior = factor(e93_sensitive_behavior)) %>% 
  # Anderson-darling test for multiple distributions
  PMCMRplus::adAllPairsTest(dist_to_motif ~ behavior, ., p.adjust.method = "fdr") 

```

```{r,include=F}
summit_nearest_e93_motif %>% 
  data.frame %>% 
  dplyr::filter(has_motif == TRUE) %>% 
  dplyr::mutate(behavior = factor(e93_sensitive_behavior)) %>% 
  # Anderson-darling test for multiple distributions
  PMCMRplus::adAllPairsTest(dist_to_motif ~ behavior, ., p.adjust.method = "fdr") -> x

# Idea: show p-val signif stars on figure legend for ecdf plot?

# Below is spec code for parsing PMCMRplus output for adding to ggplot annotation
# use levels(x$model$g) to assign levels to c1 & c2 below

# for ECDF plot: consider taking plot as input, then use cowplot to grab legend, etc.
# for boxplot: grab max values from x$model$x and adjust lines from there, return annotation df
#  user will have to configure y positions
x$p.value %>% 
  data.frame %>% 
  tibble::rownames_to_column('v') %>% 
  split(.$v) %>% 
  purrr::imap(~{
    .x %>% 
      dplyr::select(-v) %>% 
      dplyr::rename_all(function(x){paste0(x, "_", .y)})
  }) %>% 
  dplyr::bind_cols() %>% 
  tidyr::pivot_longer(dplyr::everything(), names_to = "vs", values_to = "pval") %>% 
  tidyr::separate(vs, c("c1", "c2")) %>% 
  dplyr::filter(!is.na(pval)) %>% 
  dplyr::mutate(c1 = factor(c1, levels = levels(x$model$g)),
                c2 = factor(c2, levels = levels(x$model$g)))
```


# Below is in development

```{r}
# remove when done
knitr::opts_chunk$set(eval = F)
```

## AME analysis by binding category (using dreme_results as query??)

```{r ame_by_binding, cache=TRUE, cache.comments=FALSE}
ame_by_binding <- by_binding %>% 
  purrr::map(~{
    runAme(.x, database = list(dreme_results))
  }) %>% 
  dplyr::bind_rows(.id = "binding_type") 
```

Plot heatmap of results by binding type.
```{r, dependson=ame_by_binding, fig.height=4, fig.width=10}
ame_by_binding %>% 
  ame_plot_heatmap(group = binding_type)
```

```{r}
denovo_motifs <- all_res %>% 
  dplyr::mutate(id = paste0("denovo_", best_match_id)) %>% 
  update_motifs()
```


## Denovo motif similarity by binding
```{r dreme_by_binding, cache=TRUE, cache.comments=FALSE}
dreme_by_binding <- by_binding %>% 
  purrr::map(~{
    runDreme(.x, "shuffle")
  }) %>% 
  dplyr::bind_rows(.id = "binding_type")
```

Rename the motifs to indicate the binding category they were discovered in.
```{r}
dreme_by_binding_renamed <- dreme_by_binding %>% 
  dplyr::mutate(id = paste(binding_type, seq, sep = "_")) %>% 
  # update_motifs updates the information in the special `motif` column
  update_motifs()
```


```{r, fig.height=7, fig.width=10}
cols <- colorRampPalette(c("white", "dodgerblue4"))(255)

anno.df <- dreme_by_binding_renamed %>% 
  dplyr::select(id, binding_type) %>% 
  tibble::column_to_rownames("id")

dreme_by_binding_renamed %>% 
  {universalmotif::compare_motifs(.$motif, method = "PCC")} %>% 
  pheatmap::pheatmap(color = cols, 
                     breaks = seq(0, 1, by = 1/255), 
                     annotation_col = anno.df, 
                     show_colnames = FALSE) 
```

